Quality Analyis of White Wines by Chemical Make-up by Joe Amlung
================================================================

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr)
library(scales)
library(GGally)
library(plyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
winedf <- read.csv('wineQualityWhites.csv')
```

> This is an expoloratory report interested in the quality of different white wines based on their chemical make-up. The data was published by Cortez, et. al. in 2009 in their paper Wine Quality.

# Univariate Plots Section

> In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}
summary(winedf)
```

> The summary data shows a much larger range of of values for many of the variables than I would expect. For example, the minimum sulfur dioxide is 9 (mg / dm^3) and the max is 440 (mg / dm^3), I would of initally expected the range to be much smaller. 

```{r echo = FALSE, Univarite_Plots}
str(winedf)
```

```{r echo = FALSE, Univarite_Plots}
fixed_acidity <- ggplot(aes(x = fixed.acidity), data = winedf) + 
  geom_histogram()
volatile_acidity <- ggplot(aes(x = volatile.acidity), data = winedf) + 
  geom_histogram()
citric_acid <- ggplot(aes(x = citric.acid), data = winedf) + 
  geom_histogram()
residual_sugar <- ggplot(aes(x = residual.sugar), data = winedf) + 
  geom_histogram()
chlorides <- ggplot(aes(x = chlorides), data = winedf) + 
  geom_histogram()
free_sulfur <- ggplot(aes(x = free.sulfur.dioxide), data = winedf) + 
  geom_histogram()
total_sulfur <- ggplot(aes(x = total.sulfur.dioxide), data = winedf) + 
  geom_histogram()
pH <- ggplot(aes(x = pH), data = winedf) + 
  geom_histogram()
density <- ggplot(aes(x = density), data = winedf) + 
  geom_histogram()
sulphates <- ggplot(aes(x = sulphates), data = winedf) + 
  geom_histogram()
alcohol <- ggplot(aes(x = alcohol), data = winedf) + 
  geom_histogram()
quality <- ggplot(aes(x = quality), data = winedf) + 
  geom_histogram() 
grid.arrange(fixed_acidity, volatile_acidity, citric_acid, residual_sugar, chlorides, free_sulfur, total_sulfur, pH, density, sulphates, alcohol, quality)

```

> The data, in general, has a fairly normal distrubution amongst the varibables, the noted exception being alcohol content. There are some interesting points in the intial graphs through: there is a large spike in residual sugar and sulphates looks like it could have a potential bi-modal distribution.

```{r echo = FALSE, Univarite_Plots}
ggplot(aes(x = residual.sugar), data = winedf) +
  geom_histogram(binwidth = 0.5, color = 'black', fill = '#099DD9') + 
  scale_x_continuous(breaks = seq(0,30,1), limits = c(0,30))+
  ggtitle('Count of Residual Sugar Levels', )
```

> Playing around with the bin widths, the residual sugar content looks like it is actually long-tailed.

```{r echo  FALSE, Univarite_plots}
ggplot(aes(x = alcohol), data = winedf) +
  geom_histogram(binwidth = 0.05, color = 'black', fill = '#099DD9') + 
  scale_x_sqrt() +
  ggtitle('Count of the Square Root of Alchol Content Levels') +
  labs(x = 'Total Alcohol Content')
``` 
> if we transform the alcohol data by taking the square root, we start to see that it also has some long-tail characteristics.

```{r echo = FALSE Univariate_plots}
ggplot(aes(x = sulphates), data = winedf) +
  geom_histogram(binwidth = 0.01, color = 'black', fill = '#099DD9') 
```

> Changing the binwidth allows us to see that sulphate content actually has a fairly normal distribution as well, albeit with some outliers

> To make the quality data easier to work with, I'm going to split it into 4 categories bases on the quarter percentiles. It should be noted that the distribution isn't going to be perfect since the bulk of the data has a rating of 6. Another significant digit in quality would of helped this issue significatantly. 

```{r echo = True, Univariate_plots}

cond <- winedf$quality < 5
winedf$quality_class <- ifelse(cond, 'Poor', 'High')
cond <- winedf$quality == 5
winedf$quality_class <- ifelse(cond, 'Medium-low', winedf$quality_class)
cond <- winedf$quality == 6
winedf$quality_class <- ifelse(cond, 'Medium-high', winedf$quality_class)
```

> Now that we can facet wrap or color the variables with the quality class. I'm interested to know if the large spike in the residual sugar plot is from a specific class or not. Here is the same plot shown earlier but with fill set to quality_class

```{r echo = FALSE, Univariate_plots}
ggplot(aes(x = residual.sugar, fill = quality_class), data = winedf) + 
  geom_histogram(binwidth = 0.5, color = 'black') + 
  scale_x_continuous(breaks = seq(0,30,1), limits = c(0,30)) +
  ggtitle('Count of Residual Sugar Levels by Quality_Class') +
  scale_color_brewer(type = 'qual')
```

>Looks like my hunch of the spike in residual sugar being tied to quality was wrong, looks lie that a variable indepdent of quality. 


# Univariate Analysis

The data set is orginally structured with 4898 observations of 12 variables. 11 of those variables are chemical properties related to a specific white wine. One of the more interesting things about this data set is the conformity and the shape of the distribution.

For the most part, there are actually very few outliers. Most of the properties are normally distributed and the tails cut off relatively quickly. This is probably due to wine being unpalletable with extreme chemical properties or it changes the classfication of the wine (white to red for example). There were a few noted properties that do not take on a normal distribution: sugar and alcohol content both appear to be long tailed. To obtain the tal for the alcohol content though, a square root transformation was performed to reduce some of the larger spikes in the data set.

My guess is that due to the relative uniformity of the data, that a potential accurate model should be able to be constructed to determine the quality rating of the wine. However, I have heard that wine tastings are very subjective and not scientific, so it will be interesting to see if quality is something that can be modeled. Another thing that would be intresting to check out is how much some of the chemical properties relate to eachother. For example, as someone with a chemistry background, I know that larger acid values should lead to a lower pH so I'm curious if that is indeed true or if something is "soaking? them up, leaving the pH unaffected. 

A new variable was created, quality_class, which will allow easier interpretation as to whether other variables are having some effect on the quality rating of the wine. 

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!